---
title: "Dorado Misassignment Analysis — Both-end"
output: html_document
---

```{r setup, include=FALSE, dev='ragg_png'}
knitr::opts_chunk$set(dev = "ragg_png", error = FALSE)

suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
})
```

```{r}
input_file <- knitr::current_input(dir = TRUE)
PROJECT_DIR <- if (length(input_file) == 1L && nzchar(input_file)) {
  dirname(normalizePath(input_file, mustWork = TRUE))
} else {
  normalizePath(getwd(), mustWork = TRUE)
}

SUMMARY_ROOT <- Sys.getenv(
  "DORADO_RESULTS_ROOT",
  unset = file.path(dirname(PROJECT_DIR), "data", "dorado_results")
)

DORADO_MODE <- "both_ends"
OUTPUT_SLUG <- "both_end"

RUN_DIRS <- c(
  "20251018_MPI_CLO_3ng",
  "20251019_LCI_PMO_3ng",
  "20251020_MLI_PCO_20ng",
  "20251023_CPI_LMO_20ng",
  "20251026_PMI_LCO_20ng",
  "20251029_LMI_CPO_3ng"
)

SUMMARY_PATHS <- file.path(
  SUMMARY_ROOT,
  RUN_DIRS,
  DORADO_MODE,
  "merged_bam"
)

SAMPLESHEET_PATHS <- file.path(PROJECT_DIR, "sample_sheet")

OUTDIR <- file.path(PROJECT_DIR, "dorado_species_qc_out", OUTPUT_SLUG)

dir.create(OUTDIR, showWarnings = FALSE, recursive = TRUE)
dir.create(file.path(OUTDIR, "plots"), showWarnings = FALSE, recursive = TRUE)

na_strings <- c("NA","na","N/A","n/a","","null","NULL")
```


```{r}
method_map <- data.table::fread(text="
batch_dir\tsample\tmethod
20251018_MPI_CLO\tM\tSFB
20251018_MPI_CLO\tP\tSFB
20251018_MPI_CLO\tL\tPLP
20251018_MPI_CLO\tC\tPLP
20251018_MPI_CLO\tI\tSFB
20251018_MPI_CLO\tO\tPLP
20251019_LCI_PMO_3ng\tL\tSFB
20251019_LCI_PMO_3ng\tC\tSFB
20251019_LCI_PMO_3ng\tP\tSFB+PLP
20251019_LCI_PMO_3ng\tM\tSFB+PLP
20251019_LCI_PMO_3ng\tI\tSFB
20251019_LCI_PMO_3ng\tO\tSFB+PLP
20251029_LMI_CPO_3ng\tL\tStandard
20251029_LMI_CPO_3ng\tC\tPLP
20251029_LMI_CPO_3ng\tP\tPLP
20251029_LMI_CPO_3ng\tM\tStandard
20251029_LMI_CPO_3ng\tI\tStandard
20251029_LMI_CPO_3ng\tO\tPLP
20251020_MLI_PCO_20ng\tM\tSFB
20251020_MLI_PCO_20ng\tL\tSFB
20251020_MLI_PCO_20ng\tP\tSFB+PLP
20251020_MLI_PCO_20ng\tC\tSFB+PLP
20251020_MLI_PCO_20ng\tI\tSFB
20251020_MLI_PCO_20ng\tO\tSFB+PLP
20251023_CPI_LMO_20ng\tC\tSFB
20251023_CPI_LMO_20ng\tP\tSFB
20251023_CPI_LMO_20ng\tL\tPLP
20251023_CPI_LMO_20ng\tM\tPLP
20251023_CPI_LMO_20ng\tI\tSFB
20251023_CPI_LMO_20ng\tO\tPLP
20251026_PMI_LCO_20ng\tM\tStandard
20251026_PMI_LCO_20ng\tP\tStandard
20251026_PMI_LCO_20ng\tL\tPLP
20251026_PMI_LCO_20ng\tC\tPLP
20251026_PMI_LCO_20ng\tI\tStandard
20251026_PMI_LCO_20ng\tO\tPLP
", sep="\t", na.strings = na_strings)

mapping_tbl <- data.table::fread(text="
prefix\tspecies
ATCC_23114\tMycoplasma hominis
ATCC_8482\tPhocaeicola vulgatus
ATCC_6940\tCorynebacterium striatum
DCS_Lambda\tDCS_Lambda
", sep="\t", na.strings = na_strings)

expected_tbl <- data.table(
  sample = c("M","P","C","L","I","O"),
  expected_species = c("Mycoplasma hominis","Phocaeicola vulgatus","Corynebacterium striatum","DCS_Lambda", NA_character_, NA_character_)
)

setnames(method_map, c("batch_dir","sample","method"))
method_map[, batch_dir := as.character(batch_dir)]
method_map[, sample := as.character(sample)]
method_map[, method := as.character(method)]

mapping_tbl[, prefix := as.character(prefix)]
mapping_tbl[, species := as.character(species)]
```

```{r}
list_summary_files <- function(paths) {
  out <- c()
  for (p in paths) {
    if (dir.exists(p)) {
      fs <- list.files(p, pattern="sequencing_summary.*\\.txt(\\.gz)?$|sequencing_summary\\.txt(\\.gz)?$",
                       full.names = TRUE)
      out <- c(out, fs)
    } else if (file.exists(p)) {
      out <- c(out, p)
    }
  }
  unique(out)
}

list_sheet_files <- function(paths) {
  out <- c()
  for (p in paths) {
    if (dir.exists(p)) {
      fs <- list.files(p, pattern="\\.tsv$|\\.txt$|\\.csv$", full.names = TRUE)
      out <- c(out, fs)
    } else if (file.exists(p)) {
      out <- c(out, p)
    }
  }
  unique(out)
}

summary_files <- list_summary_files(SUMMARY_PATHS)
sheet_files   <- list_sheet_files(SAMPLESHEET_PATHS)

if (length(summary_files) == 0) {
  stop(
    paste0(
      "No sequencing_summary files found. Set DORADO_RESULTS_ROOT to the directory ",
      "containing the six run directories. Paths checked:\n  ",
      paste(SUMMARY_PATHS, collapse = "\n  ")
    )
  )
}
if (length(sheet_files) == 0) stop("No sample_sheet files found. Check SAMPLESHEET_PATHS.")

message(sprintf("Found %d sequencing_summary files", length(summary_files)))
message(sprintf("Found %d sample_sheet files", length(sheet_files)))

```

```{r}
read_sheet <- function(f) {
  ss <- fread(f, na.strings = na_strings, showProgress = FALSE)
  need <- c("sample_id","barcode","alias")
  miss <- setdiff(need, names(ss))
  if (length(miss) > 0) {
    stop(sprintf("Sample sheet %s missing columns: %s", f, paste(miss, collapse=", ")))
  }
  ss <- ss[, .(sample_id = as.character(sample_id),
               barcode = as.character(barcode),
               sample = as.character(alias))]
  ss[, sheet_file := f]
  ss
}

ss_all <- rbindlist(lapply(sheet_files, read_sheet), fill = TRUE)
ss_all <- ss_all[!is.na(sample_id) & !is.na(barcode) & barcode != ""]
dup <- ss_all[, .N, by=.(sample_id, barcode)][N > 1]
if (nrow(dup) > 0) {
  message("WARNING: duplicated (sample_id, barcode) in sample sheets. Keeping the first occurrence.")
  ss_all <- ss_all[order(sheet_file)]
  ss_all <- unique(ss_all, by=c("sample_id","barcode"))
}
colnames(ss_all)[1] <- "experiment_id"
```

```{r}
wanted <- c(
  "passes_filtering","mean_qscore_template",
  "experiment_id","alias","type",
  "barcode_score","barcode_front_score","barcode_rear_score",
  "alignment_genome"
)

read_summary <- function(f) {
  hdr <- fread(f, sep="\t", nrows=0, showProgress=FALSE)
  cols <- names(hdr)
  sel <- intersect(wanted, cols)
  if (!("experiment_id" %in% sel)) stop(sprintf("%s missing experiment_id", f))
  if (!("alignment_genome" %in% sel)) stop(sprintf("%s missing alignment_genome (need the joined summary)", f))

  dt <- fread(f, sep="\t", select = sel, na.strings = na_strings, showProgress = TRUE)
  # barcode call: prefer alias/type like "barcode41"
  if (!("alias" %in% names(dt))) dt[, alias := NA_character_]
  if (!("type"  %in% names(dt))) dt[, type  := NA_character_]

  dt[, alias := as.character(alias)]
  dt[, type  := as.character(type)]


  dt[, barcode := fifelse(
    !is.na(alias) & grepl("^barcode\\d+$", alias),
    alias,
    fifelse(
      !is.na(type) & grepl("^barcode\\d+$", type),
      type,
      NA_character_
    )
  )]
  dt[, barcode := as.character(barcode)]
  dt[, experiment_id := as.character(experiment_id)]

  # aligned?
  dt[, aligned := !(is.na(alignment_genome) | alignment_genome == "*" | alignment_genome == "")]
  # numeric columns
  num_cols <- c("mean_qscore_template",
                "barcode_score","barcode_front_score","barcode_rear_score")
  for (cn in intersect(num_cols, names(dt))) dt[, (cn) := as.numeric(get(cn))]

  dt
}

dt <- rbindlist(lapply(summary_files, read_summary), fill = TRUE)
```

```{r}
# join sample letter (M/P/C/L/I/O) using experiment_id + barcode
dt <- merge(dt, ss_all[, .(experiment_id, barcode, sample)], by=c("experiment_id","barcode"), all.x=TRUE, sort=FALSE)

# Classify every Dorado barcode result against the experiment's sample sheet.
dt[, barcode_usage := fifelse(
  is.na(barcode) | barcode == "",
  "unassigned",
  fifelse(is.na(sample) | sample == "", "unused_barcode", "used_barcode")
)]

# Reads with no Dorado barcode call are true unclassified. Reads assigned to a
# barcode with no matching sample-sheet entry are unused-barcode assigned.
dt[is.na(sample) | sample == "", sample := "unclassified"]

# Unclassified and unused-barcode reads have no library-preparation method.
dt <- merge(dt, method_map, by.x=c("experiment_id","sample"), by.y=c("batch_dir","sample"), all.x=TRUE, sort=FALSE)
dt[is.na(method), method := "Unknown"]

# Attach the expected species.
dt <- merge(dt, expected_tbl, by="sample", all.x=TRUE, sort=FALSE)

dt[, observed_species := fifelse(aligned, NA_character_, "unmapped")]
for (i in seq_len(nrow(mapping_tbl))) {
  pref <- mapping_tbl$prefix[i]
  sp   <- mapping_tbl$species[i]
  dt[aligned == TRUE & (startsWith(alignment_genome, pref) | grepl(pref, alignment_genome, fixed=TRUE)),
     observed_species := sp]
}
unrecognized_references <- unique(dt[aligned == TRUE & is.na(observed_species), alignment_genome])
if (length(unrecognized_references)) {
  stop("Unrecognized aligned reference names: ", paste(unrecognized_references, collapse = ", "))
}
```

```{r}
# Retain rows accepted by Dorado.
core <- dt[passes_filtering == TRUE]
rm(dt)
invisible(gc())

# Defined-DNA reads are correct, misassigned, or unaligned. For blanks, every
# mapped read is misassigned. Reads without a sample have no sample status.
core[, status_sample := fifelse(
  sample %in% c("M","P","C","L","I","O"),

  fifelse(
    sample %in% c("I","O"),
    fifelse(
      observed_species == "unmapped",
      "unaligned",
      "misassigned"
    ),

    fifelse(
      observed_species == expected_species,
      "correct",
      fifelse(observed_species == "unmapped", "unaligned", "misassigned")
    )
  ),

  NA_character_
)]

```

```{r}
# Tag input amount from experiment_id suffix
core[, input_ng := fifelse(grepl("3ng$", experiment_id), "3 ng",
                    fifelse(grepl("20ng$", experiment_id), "20 ng", NA_character_))]

core[experiment_id == "20251018_MPI_CLO", input_ng := "3 ng"] 

core[, input_ng := factor(input_ng, levels = c("3 ng", "20 ng"))]

# All rates use every read in the merged summary for that batch. Mapped and
# unmapped are subdivided only for reads assigned to a used sample-sheet
# barcode; no-barcode and unused-barcode reads remain separate categories.
experiment_methods <- method_map[
  , .(methods = paste(unique(method), collapse = " vs ")),
  by = .(experiment_id = batch_dir)
]

batch_read_composition <- core[
  , .(
    total_reads = .N,
    classified_reads = sum(barcode_usage == "used_barcode"),
    unclassified_reads = sum(barcode_usage == "unassigned"),
    unused_barcode_reads = sum(barcode_usage == "unused_barcode"),
    classified_mapped_reads = sum(barcode_usage == "used_barcode" & aligned == TRUE),
    classified_unmapped_reads = sum(barcode_usage == "used_barcode" & aligned == FALSE)
  ),
  by = .(experiment_id, input_ng)
]

batch_read_composition <- merge(
  batch_read_composition,
  experiment_methods,
  by = "experiment_id",
  all.x = TRUE,
  sort = FALSE
)
setcolorder(batch_read_composition, c(
  "experiment_id", "input_ng", "methods", "total_reads",
  "classified_reads", "unclassified_reads", "unused_barcode_reads",
  "classified_mapped_reads", "classified_unmapped_reads"
))
setorder(batch_read_composition, experiment_id)

batch_read_composition_pct <- batch_read_composition[
  , .(
    experiment_id,
    input_ng,
    methods,
    classified_pct = 100 * classified_reads / total_reads,
    unclassified_pct = 100 * unclassified_reads / total_reads,
    unused_barcode_pct = 100 * unused_barcode_reads / total_reads,
    classified_mapped_pct = 100 * classified_mapped_reads / total_reads,
    classified_unmapped_pct = 100 * classified_unmapped_reads / total_reads
  )
]
pct_cols <- grep("_pct$", names(batch_read_composition_pct), value = TRUE)
batch_read_composition_pct[
  , (pct_cols) := lapply(.SD, round, digits = 4),
  .SDcols = pct_cols
]

stopifnot(all(
  batch_read_composition$classified_reads +
    batch_read_composition$unclassified_reads +
    batch_read_composition$unused_barcode_reads ==
    batch_read_composition$total_reads
))
stopifnot(all(
  batch_read_composition$classified_mapped_reads +
    batch_read_composition$classified_unmapped_reads ==
    batch_read_composition$classified_reads
))

fwrite(
  batch_read_composition,
  file.path(OUTDIR, "read_composition_counts_by_batch.tsv"),
  sep = "\t"
)
fwrite(
  batch_read_composition_pct,
  file.path(OUTDIR, "read_composition_percentages_by_batch.tsv"),
  sep = "\t"
)

# Start from the complete sample-sheet/method grid so a zero-read barcode would
# still contribute. Each batch contains two methods and three barcodes/method.
experiment_input <- unique(core[, .(experiment_id, input_ng)])
barcode_instance_grid <- merge(
  method_map[, .(experiment_id = batch_dir, method, sample)],
  experiment_input,
  by = "experiment_id",
  all.x = TRUE,
  sort = FALSE
)

barcode_instance_counts <- core[
  barcode_usage == "used_barcode",
  .(total_reads = .N),
  by = .(experiment_id, sample)
]

barcode_instance_counts <- merge(
  barcode_instance_grid,
  barcode_instance_counts,
  by = c("experiment_id", "sample"),
  all.x = TRUE,
  sort = FALSE
)
barcode_instance_counts[is.na(total_reads), total_reads := 0L]

reads_per_barcode_by_batch <- barcode_instance_counts[
  , .(reads_per_barcode = round(mean(total_reads), digits = 2)),
  by = .(experiment_id, input_ng, method)
]
stopifnot(all(barcode_instance_counts[, .N, by = .(experiment_id, method)]$N == 3L))
stopifnot(all(reads_per_barcode_by_batch[, .N, by = experiment_id]$N == 2L))
setorder(reads_per_barcode_by_batch, experiment_id, method)

fwrite(
  reads_per_barcode_by_batch,
  file.path(OUTDIR, "reads_per_barcode_by_batch_method.tsv"),
  sep = "\t"
)

```

### Read-category definitions

Downstream summaries include rows with `passes_filtering == TRUE` in the merged
Dorado summaries.

- **Classified**: Dorado assigned a barcode present in that experiment's sample
  sheet. These reads can be assigned to a sample and protocol.
- **True unclassified**: Dorado made no barcode call. These reads have no sample
  or protocol assignment.
- **Unused-barcode assigned**: Dorado called a barcode that is absent from that
  experiment's sample sheet. These reads also have no sample or protocol
  assignment, but remain distinct from true unclassified reads.
- **Mapped**: aligned to one of the four genomes in the Dorado reference FASTA.
- **Unmapped**: no alignment to those genomes.

For Defined-DNA samples, `correct` means mapped to the expected genome,
`misassigned` means mapped to a different target genome, and `unaligned` means
no alignment. For water samples, any mapped read is `misassigned` and a read
without an alignment is `unaligned`. True unclassified and unused-barcode
assigned reads cannot be matched to a sample or protocol, so they do not receive
one of these sample-level statuses. Dorado alignment is nevertheless attempted
for all reads, including true unclassified and unused-barcode assigned reads.

### Table: read classification and mapping composition by batch — read counts

The denominator is every read in the merged summary for that batch. Classified,
true unclassified, and unused-barcode assigned counts partition that denominator.
Mapped and unmapped are reported only for classified reads; true unclassified
and unused-barcode assigned reads are not subdivided by mapping status here.

```{r}
knitr::kable(batch_read_composition)
```

### Table: read classification and mapping composition by batch — percentages

Every percentage uses all reads in that batch's merged summary as the denominator. Therefore,
classified mapped and classified unmapped percentages sum to the classified
percentage.

```{r}
knitr::kable(batch_read_composition_pct, digits = 4)
```

### Table: reads per barcode by batch and method

Each batch is kept separate. Reads per barcode is the mean number of reads
assigned to each of the three sample-sheet barcodes belonging to that method in
that batch. Both mapped and unmapped reads are included; true unclassified and
unused-barcode assigned reads are excluded.

```{r}
reads_per_barcode_display <- copy(reads_per_barcode_by_batch)
reads_per_barcode_display[, input_ng := as.character(input_ng)]
reads_per_barcode_display[
  duplicated(experiment_id),
  `:=`(experiment_id = "", input_ng = "")
]
knitr::kable(reads_per_barcode_display, digits = 2)
```

### Performance summary

- Confusion-matrix rows are used sample-sheet barcodes. Columns are the four
  target genomes plus `Unmapped`.
- `Unmapped` contains reads without an alignment to the reference genomes.
- Blank barcodes are retained so target-mapped reads assigned to blanks
  contribute to recall denominators.
- Per-barcode precision, recall, and F1 are computed for Defined-DNA barcodes only.
- Precision is the expected-species count divided by all target-mapped reads
  assigned to that Defined-DNA barcode. Recall is the same expected-species count
  divided by all reads in the batch assigned to any used barcode and mapped to
  that expected species.
- The final performance summary reports the mean recall, precision, and F1
  across all Defined-DNA barcode instances within each method.
- True unclassified, unused-barcode assigned, and unmapped reads do not enter
  precision, recall, or F1.


```{r}
defined_dna_samples <- c("M","P","C","L")
blank_samples <- c("I","O")

target_species <- unique(na.omit(as.character(
  expected_tbl[sample %in% defined_dna_samples, expected_species]
)))
sp_cols <- target_species
target_species_ext <- c(target_species, "Unmapped")

# Build the confusion matrix, including blank barcodes.
performance_reads <- core[
  method != "Unknown" &
    sample %in% c(defined_dna_samples, blank_samples)
]

performance_reads[, observed_species_performance := ifelse(
  !is.na(observed_species) & observed_species %in% target_species,
  as.character(observed_species),
  "Unmapped"
)]

counts_long <- performance_reads[
  , .(N = .N),
  by = .(experiment_id, method, sample, observed_species = observed_species_performance)
]

grid_rows <- unique(method_map[
  method != "Unknown" & sample %in% c(defined_dna_samples, blank_samples),
  .(experiment_id = batch_dir, method, sample)
])

grid_full <- grid_rows[
  , .(observed_species = target_species_ext),
  by = .(experiment_id, method, sample)
]

performance_counts <- merge(
  grid_full, counts_long,
  by = c("experiment_id","method","sample","observed_species"),
  all.x = TRUE
)
performance_counts[is.na(N), N := 0L]

confusion_wide <- dcast(
  performance_counts,
  experiment_id + method + sample ~ observed_species,
  value.var = "N",
  fill = 0
)

fwrite(
  confusion_wide,
  file.path(OUTDIR, "performance_confusion_counts.tsv"),
  sep = "\t"
)

# Per-Defined-DNA-barcode precision, recall, and F1.
exp_map <- unique(
  expected_tbl[sample %in% defined_dna_samples, .(sample, expected_species)]
)

tab_g <- merge(
  confusion_wide[sample %in% defined_dna_samples],
  exp_map,
  by = "sample",
  all.x = TRUE
)

# Performance metrics are conditional on a read being assigned to a used
# barcode and mapped to a target species. True unclassified, unused-barcode
# assigned, and unmapped reads do not enter the metrics.
col_tot <- confusion_wide[
  , lapply(.SD, sum),
  by = .(experiment_id),
  .SDcols = sp_cols
]

col_tot_long <- melt(
  col_tot,
  id.vars = "experiment_id",
  variable.name = "expected_species",
  value.name = "col_total_expected"
)

tab_g[, row_total_species := rowSums(as.matrix(.SD)), .SDcols = sp_cols]

m_g   <- as.matrix(tab_g[, ..sp_cols])
idx_c <- match(tab_g$expected_species, sp_cols)

tp <- rep(NA_real_, nrow(tab_g))
ok <- which(!is.na(idx_c))
if (length(ok)) tp[ok] <- m_g[cbind(ok, idx_c[ok])]
tab_g[, tp := tp]

tab_g <- merge(
  tab_g,
  col_tot_long,
  by = c("experiment_id","expected_species"),
  all.x = TRUE
)

tab_g[, precision := fifelse(row_total_species > 0, tp / row_total_species, NA_real_)]
tab_g[, recall    := fifelse(col_total_expected > 0, tp / col_total_expected, NA_real_)]
tab_g[, f1 := fifelse(
  is.finite(precision) & is.finite(recall) & (precision + recall) > 0,
  2 * precision * recall / (precision + recall),
  NA_real_
)]

fwrite(
  tab_g[, .(
    experiment_id, method, sample, expected_species,
    tp, row_total_species, col_total_expected,
    precision, recall, f1
  )],
  file.path(OUTDIR, "performance_barcode_instance_metrics.tsv"),
  sep = "\t"
)

# Method-level performance summary.
performance_summary <- tab_g[, .(
  Mean_Recall_pct    = 100 * mean(recall,    na.rm = TRUE),
  Mean_Precision_pct = 100 * mean(precision, na.rm = TRUE),
  Mean_F1_pct        = 100 * mean(f1,        na.rm = TRUE),
  n_defined_dna_instances = .N,
  n_batches          = uniqueN(experiment_id)
), by = .(Method = method)]

setorder(performance_summary, Method)

fwrite(
  performance_summary,
  file.path(OUTDIR, "performance_summary_by_method.tsv"),
  sep = "\t"
)

if (requireNamespace("knitr", quietly = TRUE)) {
  knitr::kable(performance_summary, digits = 3)
} else {
  performance_summary
}
```


### Summary: sample-level misassigned per million mapped reads with shared denominator

- The denominator is the total number of mapped reads assigned to the
  used barcodes of one method within one batch.
- True unclassified, unused-barcode assigned, and unmapped reads are excluded.
- For Defined DNA, the numerator is mapped reads assigned to that barcode but
  aligned to a different genome. For water blanks, every mapped read assigned
  to the blank barcode is counted as misassigned.

```{r}
ord <- c("Standard", "SFB", "PLP", "SFB+PLP")
sample_label_levels <- c(
  "Water",
  "Mycoplasma hominis",
  "DCS_Lambda",
  "Phocaeicola vulgatus",
  "Corynebacterium striatum"
)

# Per-sample numerator and denominator pieces
sample_metrics <- core[
  method != "Unknown" &
    sample %in% c(defined_dna_samples, blank_samples),
  .(
    total_mapped = sum(observed_species %in% target_species),
    tp = sum(
      observed_species %in% target_species &
      observed_species == expected_species
    )
  ),
  by = .(experiment_id, method, input_ng, sample, expected_species)
]

sample_metrics[, method := factor(method, levels = ord)]

sample_metrics[, sample_label := fifelse(
  sample %in% blank_samples,
  "Water",
  expected_species
)]

# For DNA samples: misassigned = target-mapped non-expected reads
# For blanks: misassigned = all target-mapped reads
sample_metrics[, misassigned_for_plot := fifelse(
  sample %in% blank_samples,
  total_mapped,
  pmax(total_mapped - tp, 0)
)]

# Shared denominator within each batch × method
denom_tbl <- sample_metrics[
  ,
  .(total_mapped_batch_method = sum(total_mapped, na.rm = TRUE)),
  by = .(experiment_id, method, input_ng)
]

plot_shared <- merge(
  sample_metrics,
  denom_tbl,
  by = c("experiment_id", "method", "input_ng"),
  all.x = TRUE,
  sort = FALSE
)

plot_shared[, misassigned_per_million_shared := fifelse(
  total_mapped_batch_method > 0,
  1e6 * misassigned_for_plot / total_mapped_batch_method,
  NA_real_
)]

fwrite(
  plot_shared,
  file.path(OUTDIR, "misassigned_per_million_mapped_by_sample.tsv"),
  sep = "\t"
)

pair_tbl <- unique(plot_shared[, .(experiment_id, method)])[
  ,
  .(batch_pair = paste(sort(unique(as.character(method))), collapse = " vs ")),
  by = experiment_id
]

df_pairs <- merge(
  plot_shared[, .(
    experiment_id, method, input_ng,
    sample_label, misassigned_per_million_shared
  )],
  pair_tbl,
  by = "experiment_id",
  all.x = TRUE,
  sort = FALSE
)

df_pairs[, sample_label := factor(sample_label, levels = sample_label_levels)]
```

### Figure: Misassigned per million mapped reads by sample (shared denominator within batch × method)

Each sample's numerator is its misassigned mapped reads. The denominator is all
correct plus misassigned mapped reads assigned to used sample-sheet barcodes
within the same batch and method. Unmapped, true unclassified, and
unused-barcode assigned reads are excluded.

```{r}
p2 <- ggplot(
  df_pairs[!is.na(batch_pair)],
  aes(x = sample_label, y = misassigned_per_million_shared, fill = method)
) +
  geom_col(position = position_dodge2(width = 0.8, preserve = "single")) +
  facet_grid(batch_pair ~ input_ng, scales = "free_y") +
  labs(
    x = "Sample",
    y = "Misassigned reads per million mapped reads\n(shared denominator within batch × method)",
    fill = "Method"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    text = element_text(size = 10),
    strip.background = element_rect(colour = NA)
  )
```

```{r fig.height=10,fig.width=8}
p2
```

### Figure: Mean sample-level misassigned per million mapped reads (Water vs Defined DNA)

Bars are means of the sample-level values above across batches and input
amounts; error bars are standard errors. The numerator and denominator rules
are unchanged from the preceding figure.

```{r}
df_summary <- df_pairs[
  ,
  .(
    mean_mis = mean(misassigned_per_million_shared, na.rm = TRUE),
    sd_mis   = sd(misassigned_per_million_shared, na.rm = TRUE),
    n        = sum(!is.na(misassigned_per_million_shared))
  ),
  by = .(
    sample_group = fifelse(sample_label == "Water", "Water", "Defined DNA"),
    method
  )
]

df_summary[, se_mis := sd_mis / sqrt(n)]
df_summary[, sample_group := factor(sample_group, levels = c("Water", "Defined DNA"))]

p1 <- ggplot(df_summary, aes(x = sample_group, y = mean_mis, fill = method)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(
    aes(ymin = mean_mis - se_mis, ymax = mean_mis + se_mis),
    position = position_dodge(width = 0.8),
    width = 0.2
  ) +
  labs(
    x = NULL,
    y = "Misassigned reads per million mapped reads",
    fill = "Protocol"
  ) +
  scale_x_discrete(labels = c(
    "Defined DNA" = "Defined DNA",
    "Water" = "Water blanks"
  )) +
  theme_bw() +
  theme(
    text = element_text(size = 12),
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )
```

```{r fig.height=5,fig.width=7}
p1
```


### Quality-score distributions by assignment class

The first four columns contain reads assigned to used sample-sheet barcodes and
show Correct, Misassigned, and Unmapped groups for Standard, SFB, PLP, and
SFB+PLP. The fifth column contains all true unclassified reads and all
unused-barcode assigned reads pooled across batches. These reads have no sample
or protocol assignment and therefore are not divided into Correct,
Misassigned, and Unmapped groups. Dorado provides no barcode score for true
unclassified reads, so that group appears only for template mean Q-score;
unused-barcode assigned reads retain their barcode scores. Each plotted group
is capped at 200,000 reads.

```{r quality-score-data, include=FALSE}
# Assignment classes used for the complete audit and the protocol-level plot
PLOT_CAP_PER_GROUP <- 200000L
set.seed(1)

qs_metrics <- intersect(
  c("mean_qscore_template", "barcode_score", "barcode_front_score", "barcode_rear_score"),
  names(core)
)

quality_cols <- c("experiment_id", "method", "barcode_usage", "status_sample", qs_metrics)
qcore <- core[, ..quality_cols]

qcore[, quality_group := fcase(
  barcode_usage == "unused_barcode", "Unused barcode",
  barcode_usage == "unassigned", "Unclassified",
  barcode_usage == "used_barcode" & status_sample == "correct", "Correct assigned",
  barcode_usage == "used_barcode" & status_sample == "misassigned", "Misassigned",
  barcode_usage == "used_barcode" & status_sample == "unaligned", "Unmapped",
  default = NA_character_
)]

qcore <- qcore[!is.na(quality_group)]

qcore[, plot_method := fifelse(
  method == "Unknown",
  "Unclassified / unused barcode",
  as.character(method)
)]

quality_group_levels <- c(
  "Correct assigned", "Misassigned", "Unmapped",
  "Unclassified", "Unused barcode"
)
plot_method_levels <- c(ord, "Unclassified / unused barcode")
qcore[, quality_group := factor(quality_group, levels = quality_group_levels)]
qcore[, plot_method := factor(plot_method, levels = plot_method_levels)]

# Dorado records -1 rather than NA for barcode scores when no barcode was
# assigned. Treat those sentinel values as missing in the complete audit data.
barcode_metrics <- intersect(
  c("barcode_score", "barcode_front_score", "barcode_rear_score"),
  qs_metrics
)
qcore[
  barcode_usage == "unassigned",
  (barcode_metrics) := lapply(.SD, function(x) NA_real_),
  .SDcols = barcode_metrics
]

quality_group_counts <- qcore[
  , .N,
  by = .(experiment_id, plot_method, quality_group)
]
fwrite(
  quality_group_counts,
  file.path(OUTDIR, "quality_assignment_class_counts.tsv"),
  sep = "\t"
)

# Plot the four protocol columns plus one combined column for true unclassified
# and unused-barcode assigned reads. Cap large groups for density estimation.
qcore_plot <- qcore[plot_method %in% plot_method_levels]
if (PLOT_CAP_PER_GROUP > 0) {
  qcore_plot <- qcore_plot[
    , .SD[sample.int(.N, min(.N, PLOT_CAP_PER_GROUP))],
    by = .(plot_method, quality_group)
  ]
}

quality_plot_counts <- qcore_plot[
  , .(reads_plotted = .N),
  by = .(plot_method, quality_group)
]
setorder(quality_plot_counts, plot_method, quality_group)
fwrite(
  quality_plot_counts,
  file.path(OUTDIR, "quality_plot_read_counts.tsv"),
  sep = "\t"
)

qlong <- melt(
  qcore_plot,
  id.vars = c("plot_method", "quality_group"),
  measure.vars = qs_metrics,
  variable.name = "metric",
  value.name = "value"
)
qlong <- qlong[is.finite(value)]

metric_lab <- c(
  mean_qscore_template = "Template mean Q-score",
  barcode_score        = "Barcode score",
  barcode_front_score  = "Barcode front score",
  barcode_rear_score   = "Barcode rear score"
)
qlong[, metric := factor(metric, levels = qs_metrics, labels = metric_lab[qs_metrics])]
qlong[, quality_group := factor(quality_group, levels = quality_group_levels)]
qlong[, plot_method := factor(plot_method, levels = plot_method_levels)]

# Confirm that all four metrics use the same reads within each protocol group.
# True unclassified reads have no barcode-score values, so this assertion does
# not apply to the combined fifth column.
quality_metric_count_check <- qlong[
  , .(metric_reads = .N),
  by = .(plot_method, metric, quality_group)
]
quality_metric_count_check <- merge(
  quality_metric_count_check,
  quality_plot_counts,
  by = c("plot_method", "quality_group"),
  all.x = TRUE,
  sort = FALSE
)
quality_metric_count_check_protocol <- quality_metric_count_check[
  plot_method %in% ord
]
stopifnot(all(
  quality_metric_count_check_protocol$metric_reads ==
    quality_metric_count_check_protocol$reads_plotted
))

p_quality <- ggplot(
  qlong,
  aes(
    x = quality_group,
    y = value,
    fill = quality_group,
    group = quality_group
  )
) +
  geom_violin(trim = TRUE, scale = "width") +
  geom_boxplot(width = 0.12, outlier.size = 0.2) +
  scale_x_discrete(
    drop = TRUE,
    labels = c(
      "Correct assigned" = "Correct",
      "Misassigned" = "Misassigned",
      "Unmapped" = "Unmapped",
      "Unclassified" = "Unclassified",
      "Unused barcode" = "Unused barcode"
    )
  ) +
  scale_fill_manual(
    values = c(
      "Correct assigned" = "#4DAF4A",
      "Misassigned" = "#E41A1C",
      "Unmapped" = "#999999",
      "Unclassified" = "#377EB8",
      "Unused barcode" = "#984EA3"
    ),
    drop = TRUE
  ) +
  facet_grid(metric ~ plot_method, scales = "free", switch = "y") +
  labs(
    x = "Assignment class",
    y = NULL,
    fill = "Assignment class"
  ) +
  theme_bw(base_size = 18) +
  theme(
    axis.text.x = element_text(size = 15, angle = 25, hjust = 1),
    axis.text.y = element_text(size = 15),
    axis.title.x = element_text(size = 18),
    legend.title = element_text(size = 17),
    legend.text = element_text(size = 16),
    legend.position = "top",
    strip.text.x = element_text(size = 16),
    strip.text.y = element_text(size = 16),
    strip.placement = "outside",
    strip.background = element_blank(),
    panel.spacing = grid::unit(0.8, "lines")
  )

ggsave(
  file.path(OUTDIR, "plots", "qscore_violin_assignment_classes.pdf"),
  p_quality,
  width = 17,
  height = 12
)
```

### Table: reads used in each quality-score violin

Each protocol × assignment-class group is listed once because the same reads
are used for all four metrics in the first four columns. The final combined
column lists true unclassified and unused-barcode assigned reads separately;
true unclassified reads contribute only template mean Q-score because they have
no barcode scores. Counts are after the plotting cap of 200,000 reads per group.

```{r}
knitr::kable(quality_plot_counts)
```

```{r quality-score-plot, fig.width=17, fig.height=12, warning=FALSE}
p_quality
```
