---
title: "Dorado Read-Length Distributions — Default"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  dev = "ragg_png",
  dpi = 180,
  fig.width = 16,
  fig.height = 9
)

suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(knitr)
})

input_file <- knitr::current_input(dir = TRUE)
PROJECT_DIR <- if (length(input_file) == 1L && nzchar(input_file)) {
  dirname(normalizePath(input_file, mustWork = TRUE))
} else {
  normalizePath(getwd(), mustWork = TRUE)
}
DATA_ROOT <- Sys.getenv(
  "DORADO_RESULTS_ROOT",
  file.path(dirname(PROJECT_DIR), "data", "dorado_results")
)
SAMPLE_SHEET_DIR <- file.path(PROJECT_DIR, "sample_sheet")
OUTDIR <- file.path(PROJECT_DIR, "dorado_read_length_out", "default")
PLOT_DIR <- file.path(OUTDIR, "plots")
dir.create(PLOT_DIR, recursive = TRUE, showWarnings = FALSE)

RUN_DIRS <- c(
  "20251018_MPI_CLO_3ng",
  "20251019_LCI_PMO_3ng",
  "20251020_MLI_PCO_20ng",
  "20251023_CPI_LMO_20ng",
  "20251026_PMI_LCO_20ng",
  "20251029_LMI_CPO_3ng"
)

EXPECTED_GENOME <- c(
  M = "Mycoplasma hominis",
  P = "Phocaeicola vulgatus",
  C = "Corynebacterium striatum",
  L = "DCS lambda",
  I = "blank",
  O = "blank"
)

PANEL_TITLES <- c(
  "20251018_MPI_CLO_3ng" = "3 ng: SFB vs PLP",
  "20251019_LCI_PMO_3ng" = "3 ng: SFB vs SFB+PLP",
  "20251020_MLI_PCO_20ng" = "20 ng: SFB vs SFB+PLP",
  "20251023_CPI_LMO_20ng" = "20 ng: SFB vs PLP",
  "20251026_PMI_LCO_20ng" = "20 ng: Standard vs PLP",
  "20251029_LMI_CPO_3ng" = "3 ng: Standard vs PLP"
)

BLANK_METHODS <- data.table(
  run_dir = rep(RUN_DIRS, each = 2),
  alias = rep(c("I", "O"), times = length(RUN_DIRS)),
  method = c(
    "SFB", "PLP",
    "SFB", "SFB+PLP",
    "SFB", "SFB+PLP",
    "SFB", "PLP",
    "Standard", "PLP",
    "Standard", "PLP"
  )
)

DISPLAY_MIN_BP <- 10
DISPLAY_MAX_BP <- 100000
LOG_BIN_WIDTH <- 0.05

extract_barcode <- function(alias, type) {
  alias <- as.character(alias)
  type <- as.character(type)
  ans <- rep(NA_character_, length(alias))
  alias_ok <- !is.na(alias) & grepl("^barcode[0-9]+$", alias)
  type_ok <- !is.na(type) & grepl("^barcode[0-9]+$", type)
  ans[alias_ok] <- alias[alias_ok]
  ans[is.na(ans) & type_ok] <- type[is.na(ans) & type_ok]
  ans
}

read_sample_sheet <- function(run_dir) {
  sheet_file <- file.path(
    SAMPLE_SHEET_DIR,
    paste0(run_dir, "_sample_sheet.csv")
  )
  if (!file.exists(sheet_file)) {
    stop("Missing sample sheet: ", sheet_file)
  }

  sheet <- fread(sheet_file, select = c("alias", "barcode"))
  sheet[, expected_genome := unname(EXPECTED_GENOME[as.character(alias)])]
  sheet[is.na(expected_genome) | expected_genome == "", expected_genome := "blank"]
  target_run <- run_dir
  blank_map <- BLANK_METHODS[run_dir == target_run]
  sheet[, blank_method := blank_map$method[match(alias, blank_map$alias)]]
  sheet[, panel_description := fifelse(
    expected_genome == "blank",
    paste(blank_method, "blank"),
    expected_genome
  )]
  sheet[, panel_label := paste0(barcode, "\n", panel_description)]
  sheet
}

process_batch <- function(run_dir) {
  summary_file <- file.path(
    DATA_ROOT,
    run_dir,
    "either_end",
    "merged_bam",
    "sequencing_summary_merged.txt"
  )
  if (!file.exists(summary_file)) {
    stop("Missing Dorado sequencing summary: ", summary_file)
  }

  sheet <- read_sample_sheet(run_dir)
  reads <- fread(
    summary_file,
    select = c("alias", "type", "sequence_length_template"),
    showProgress = FALSE
  )
  reads[, read_length_bp := as.numeric(sequence_length_template)]
  reads <- reads[is.finite(read_length_bp) & read_length_bp > 0]
  reads[, barcode := extract_barcode(alias, type)]

  label_lookup <- setNames(sheet$panel_label, sheet$barcode)
  reads[, panel_label := unname(label_lookup[barcode])]
  # Reads without a barcode call and reads assigned to a barcode absent from the
  # batch sample sheet are both included in the Unclassified panel.
  reads[is.na(panel_label), panel_label := "Unclassified"]

  panel_levels <- c(sheet$panel_label, "Unclassified", "All reads")
  displayed <- reads[
    read_length_bp >= DISPLAY_MIN_BP & read_length_bp <= DISPLAY_MAX_BP
  ]
  displayed[, log_bin_start :=
              floor(log10(read_length_bp) / LOG_BIN_WIDTH) * LOG_BIN_WIDTH]

  histogram_data <- rbindlist(list(
    displayed[, .(frequency = .N), by = .(panel_label, log_bin_start)],
    displayed[, .(frequency = .N), by = .(log_bin_start)][, panel_label := "All reads"]
  ), use.names = TRUE)
  histogram_data[, panel_label := factor(panel_label, levels = panel_levels)]
  histogram_data[, `:=`(
    bin_min_bp = 10^log_bin_start,
    bin_max_bp = 10^(log_bin_start + LOG_BIN_WIDTH)
  )]

  group_summary <- reads[, .(
    reads = .N,
    median_length_bp = median(read_length_bp),
    mean_length_bp = mean(read_length_bp),
    q95_length_bp = as.numeric(quantile(read_length_bp, 0.95)),
    max_length_bp = max(read_length_bp),
    display_min_bp = DISPLAY_MIN_BP,
    display_max_bp = DISPLAY_MAX_BP,
    reads_below_display_min = sum(read_length_bp < DISPLAY_MIN_BP),
    reads_above_display_max = sum(read_length_bp > DISPLAY_MAX_BP)
  ), by = panel_label]
  group_summary <- rbindlist(list(
    group_summary,
    reads[, .(
      panel_label = "All reads",
      reads = .N,
      median_length_bp = median(read_length_bp),
      mean_length_bp = mean(read_length_bp),
      q95_length_bp = as.numeric(quantile(read_length_bp, 0.95)),
      max_length_bp = max(read_length_bp),
      display_min_bp = DISPLAY_MIN_BP,
      display_max_bp = DISPLAY_MAX_BP,
      reads_below_display_min = sum(read_length_bp < DISPLAY_MIN_BP),
      reads_above_display_max = sum(read_length_bp > DISPLAY_MAX_BP)
    )]
  ), use.names = TRUE)
  group_summary[, `:=`(
    batch = run_dir,
    percent_below_display_min = 100 * reads_below_display_min / reads,
    percent_above_display_max = 100 * reads_above_display_max / reads
  )]
  setcolorder(group_summary, c(
    "batch", "panel_label", "reads", "median_length_bp",
    "mean_length_bp", "q95_length_bp", "max_length_bp",
    "display_min_bp", "display_max_bp", "reads_below_display_min",
    "percent_below_display_min", "reads_above_display_max",
    "percent_above_display_max"
  ))

  p <- ggplot(histogram_data) +
    geom_rect(
      aes(
        xmin = bin_min_bp,
        xmax = bin_max_bp,
        ymin = 0,
        ymax = frequency
      ),
      fill = "#3977A8",
      color = NA
    ) +
    facet_wrap(vars(panel_label), ncol = 4, scales = "free_y", drop = FALSE) +
    scale_x_log10(
      limits = c(DISPLAY_MIN_BP, DISPLAY_MAX_BP),
      breaks = c(10, 100, 1000, 10000, 100000),
      labels = c("10 bp", "100 bp", "1 kb", "10 kb", "100 kb"),
      expand = expansion(mult = 0)
    ) +
    scale_y_continuous(labels = scales::label_comma()) +
    labs(
      x = "Read length (log10 scale)",
      y = "Frequency"
    ) +
    theme_bw(base_size = 16) +
    theme(
      strip.text = element_text(size = 14, face = "bold", lineheight = 1.05),
      axis.title = element_text(size = 19),
      axis.text = element_text(size = 14),
      panel.grid.minor = element_blank(),
      plot.margin = margin(8, 12, 8, 8)
    )

  list(plot = p, summary = group_summary)
}
```

This report uses the Default Dorado barcode assignments. Each batch is shown as
six sample-sheet barcodes, one `Unclassified` group, and one `All reads` group.
Barcode labels combine the original barcode number with its expected genome;
barcodes for the two blank samples are labeled `blank`. `Unclassified` includes
both reads without a barcode call and reads assigned to a barcode not present in
that batch's sample sheet.

Histograms use equal-width 0.05-log10 bins. All panels in all six batches use the
same log10-scaled 10 bp--100 kb read-length range so that distributions can be
compared directly. The frequency axis is allowed to vary by barcode panel so
that distribution shapes remain visible. Reads outside the displayed range are
retained in the tabulated totals.

```{r distributions, results='asis', fig.width=16, fig.height=9, out.width='100%'}
all_summaries <- vector("list", length(RUN_DIRS))

for (i in seq_along(RUN_DIRS)) {
  run_dir <- RUN_DIRS[i]
  result <- process_batch(run_dir)
  all_summaries[[i]] <- result$summary

  cat("\n\n## ", unname(PANEL_TITLES[run_dir]), "\n\n", sep = "")
  print(result$plot)

  ggsave(
    filename = file.path(
      PLOT_DIR,
      paste0("read_length_distribution_", run_dir, ".pdf")
    ),
    plot = result$plot,
    width = 16,
    height = 9,
    units = "in",
    device = cairo_pdf
  )
}

summary_table <- rbindlist(all_summaries, use.names = TRUE)
fwrite(
  summary_table,
  file.path(OUTDIR, "read_length_summary_by_batch_and_barcode.tsv"),
  sep = "\t"
)
```

## Reads outside the plotted range

```{r outside-range}
outside_range <- summary_table[panel_label == "All reads", .(
  batch,
  reads_below_10_bp = reads_below_display_min,
  percent_below_10_bp = percent_below_display_min,
  reads_above_100_kb = reads_above_display_max,
  percent_above_100_kb = percent_above_display_max
)]
kable(outside_range, digits = 3, format.args = list(big.mark = ","))
```

## Read-count check

For each batch, the six sample-sheet barcode groups plus `Unclassified` partition
all reads with a positive Dorado read length. The `All reads` row is the same set
of reads pooled together and is not an additional category.

```{r count-check}
count_check <- summary_table[
  panel_label != "All reads",
  .(sum_of_seven_groups = sum(reads)),
  by = batch
]
count_check <- merge(
  count_check,
  summary_table[panel_label == "All reads", .(batch, all_reads = reads)],
  by = "batch"
)
count_check[, counts_match := sum_of_seven_groups == all_reads]
kable(count_check, format.args = list(big.mark = ","))
```
