---
title: "Untitled"
output: html_document
date: "2026-02-12"
---

```{r}
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
})
```

```{r}
SUMMARY_PATHS <- c(
  "/work/qd33/data/basecalling_data/20251018_MPI_CLO_3ng_dorado131_trim_new/merged_bam",  
  "/work/qd33/data/basecalling_data/20251019_LCI_PMO_3ng_dorado131_trim_new/merged_bam",
  "/work/qd33/data/basecalling_data/20251020_MLI_PCO_20ng_dorado131_trim_new/merged_bam",
  "/work/qd33/data/basecalling_data/20251023_CPI_LMO_20ng_dorado131_trim_new/merged_bam",
  "/work/qd33/data/basecalling_data/20251026_PMI_LCO_20ng_dorado131_trim_new/merged_bam",
  "/work/qd33/data/basecalling_data/20251029_LMI_CPO_3ng_dorado131_trim_new/merged_bam"
)

SAMPLESHEET_PATHS <- c(
  "/hpc/home/qd33/projects/protocol_modification/sample_sheet"         
)

OUTDIR <- "/hpc/home/qd33/projects/protocol_modification/dorado_species_qc_out"

MIN_MAPQ <- 0      

dir.create(OUTDIR, showWarnings = FALSE, recursive = TRUE)
dir.create(file.path(OUTDIR, "plots"), showWarnings = FALSE, recursive = TRUE)

na_strings <- c("NA","na","N/A","n/a","","null","NULL")
```


```{r}
method_map <- data.table::fread(text="
batch_dir\tsample\tmethod
20251018_MPI_CLO\tM\tSFB
20251018_MPI_CLO\tP\tSFB
20251018_MPI_CLO\tL\tPLP
20251018_MPI_CLO\tC\tPLP
20251018_MPI_CLO\tI\tSFB
20251018_MPI_CLO\tO\tPLP
20251019_LCI_PMO_3ng\tL\tSFB
20251019_LCI_PMO_3ng\tC\tSFB
20251019_LCI_PMO_3ng\tP\tSFB+PLP
20251019_LCI_PMO_3ng\tM\tSFB+PLP
20251019_LCI_PMO_3ng\tI\tSFB
20251019_LCI_PMO_3ng\tO\tSFB+PLP
20251029_LMI_CPO_3ng\tL\tStandard
20251029_LMI_CPO_3ng\tC\tPLP
20251029_LMI_CPO_3ng\tP\tPLP
20251029_LMI_CPO_3ng\tM\tStandard
20251029_LMI_CPO_3ng\tI\tStandard
20251029_LMI_CPO_3ng\tO\tPLP
20251020_MLI_PCO_20ng\tM\tSFB
20251020_MLI_PCO_20ng\tL\tSFB
20251020_MLI_PCO_20ng\tP\tSFB+PLP
20251020_MLI_PCO_20ng\tC\tSFB+PLP
20251020_MLI_PCO_20ng\tI\tSFB
20251020_MLI_PCO_20ng\tO\tSFB+PLP
20251023_CPI_LMO_20ng\tC\tSFB
20251023_CPI_LMO_20ng\tP\tSFB
20251023_CPI_LMO_20ng\tL\tPLP
20251023_CPI_LMO_20ng\tM\tPLP
20251023_CPI_LMO_20ng\tI\tSFB
20251023_CPI_LMO_20ng\tO\tPLP
20251026_PMI_LCO_20ng\tM\tStandard
20251026_PMI_LCO_20ng\tP\tStandard
20251026_PMI_LCO_20ng\tL\tPLP
20251026_PMI_LCO_20ng\tC\tPLP
20251026_PMI_LCO_20ng\tI\tStandard
20251026_PMI_LCO_20ng\tO\tPLP
", sep="\t", na.strings = na_strings)

mapping_tbl <- data.table::fread(text="
prefix\tspecies
ATCC_23114\tMycoplasma hominis
ATCC_8482\tPhocaeicola vulgatus
ATCC_6940\tCorynebacterium striatum
DCS_Lambda\tDCS_Lambda
unmapped\tunmapped
", sep="\t", na.strings = na_strings)

expected_tbl <- data.table(
  sample = c("M","P","C","L","I","O"),
  expected_species = c("Mycoplasma hominis","Phocaeicola vulgatus","Corynebacterium striatum","DCS_Lambda", NA_character_, NA_character_),
  is_blank = c(FALSE,FALSE,FALSE,FALSE,TRUE,TRUE)
)

setnames(method_map, c("batch_dir","sample","method"))
method_map[, batch_dir := as.character(batch_dir)]
method_map[, sample := as.character(sample)]
method_map[, method := as.character(method)]

mapping_tbl[, prefix := as.character(prefix)]
mapping_tbl[, species := as.character(species)]
```

```{r}
list_summary_files <- function(paths) {
  out <- c()
  for (p in paths) {
    if (dir.exists(p)) {
      fs <- list.files(p, pattern="sequencing_summary.*\\.txt(\\.gz)?$|sequencing_summary\\.txt(\\.gz)?$",
                       full.names = TRUE)
      out <- c(out, fs)
    } else if (file.exists(p)) {
      out <- c(out, p)
    }
  }
  unique(out)
}

list_sheet_files <- function(paths) {
  out <- c()
  for (p in paths) {
    if (dir.exists(p)) {
      fs <- list.files(p, pattern="\\.tsv$|\\.txt$|\\.csv$", full.names = TRUE)
      out <- c(out, fs)
    } else if (file.exists(p)) {
      out <- c(out, p)
    }
  }
  unique(out)
}

summary_files <- list_summary_files(SUMMARY_PATHS)
sheet_files   <- list_sheet_files(SAMPLESHEET_PATHS)

if (length(summary_files) == 0) stop("No sequencing_summary files found. Check SUMMARY_PATHS.")
if (length(sheet_files) == 0) stop("No sample_sheet files found. Check SAMPLESHEET_PATHS.")

message(sprintf("Found %d sequencing_summary files", length(summary_files)))
message(sprintf("Found %d sample_sheet files", length(sheet_files)))

```

```{r}
read_sheet <- function(f) {
  ss <- fread(f, na.strings = na_strings, showProgress = FALSE)
  need <- c("sample_id","barcode","alias")
  miss <- setdiff(need, names(ss))
  if (length(miss) > 0) {
    stop(sprintf("Sample sheet %s missing columns: %s", f, paste(miss, collapse=", ")))
  }
  ss <- ss[, .(sample_id = as.character(sample_id),
               barcode = as.character(barcode),
               sample = as.character(alias))]
  ss[, sheet_file := f]
  ss
}

ss_all <- rbindlist(lapply(sheet_files, read_sheet), fill = TRUE)
ss_all <- ss_all[!is.na(sample_id) & !is.na(barcode) & barcode != ""]
dup <- ss_all[, .N, by=.(sample_id, barcode)][N > 1]
if (nrow(dup) > 0) {
  message("WARNING: duplicated (sample_id, barcode) in sample sheets. Keeping the first occurrence.")
  ss_all <- ss_all[order(sheet_file)]
  ss_all <- unique(ss_all, by=c("sample_id","barcode"))
}
colnames(ss_all)[1] <- "experiment_id"
```

```{r}
wanted <- c(
  "read_id","passes_filtering","sequence_length_template","mean_qscore_template",
  "experiment_id","alias","type",
  "barcode_score","barcode_front_score","barcode_rear_score",
  "alignment_genome","alignment_mapping_quality","alignment_identity","alignment_accuracy",
  "alignment_num_alignments","alignment_num_secondary_alignments"
)

read_summary <- function(f) {
  hdr <- fread(f, sep="\t", nrows=0, showProgress=FALSE)
  cols <- names(hdr)
  sel <- intersect(wanted, cols)
  if (!("experiment_id" %in% sel)) stop(sprintf("%s missing experiment_id", f))
  if (!("alignment_genome" %in% sel)) stop(sprintf("%s missing alignment_genome (need the joined summary)", f))

  dt <- fread(f, sep="\t", select = sel, na.strings = na_strings, showProgress = TRUE)
  dt[, file := f]

  # barcode call: prefer alias/type like "barcode41"
  if (!("alias" %in% names(dt))) dt[, alias := NA_character_]
  if (!("type"  %in% names(dt))) dt[, type  := NA_character_]

  dt[, alias := as.character(alias)]
  dt[, type  := as.character(type)]


  dt[, barcode := fifelse(
    !is.na(alias) & grepl("^barcode\\d+$", alias),
    alias,
    fifelse(
      !is.na(type) & grepl("^barcode\\d+$", type),
      type,
      NA_character_
    )
  )]
dt[, barcode := as.character(barcode)]
  dt[, experiment_id := as.character(experiment_id)]

  # aligned?
  dt[, aligned := !(is.na(alignment_genome) | alignment_genome == "*" | alignment_genome == "")]
  # numeric columns
  num_cols <- c("sequence_length_template","mean_qscore_template",
                "barcode_score","barcode_front_score","barcode_rear_score",
                "alignment_mapping_quality","alignment_identity","alignment_accuracy")
  for (cn in intersect(num_cols, names(dt))) dt[, (cn) := as.numeric(get(cn))]

  dt
}

dt <- rbindlist(lapply(summary_files, read_summary), fill = TRUE)
```

```{r}
# join sample letter (M/P/C/L/I/O) using experiment_id + barcode
dt <- merge(dt, ss_all[, .(experiment_id, barcode, sample)], by=c("experiment_id","barcode"), all.x=TRUE, sort=FALSE)

# barcodes not in sample_sheet -> unclassified group
dt[is.na(sample) | sample == "", sample := "unclassified"]

# attach method: only for samples that exist in method_map; otherwise Unknown
dt <- merge(dt, method_map, by.x=c("experiment_id","sample"), by.y=c("batch_dir","sample"), all.x=TRUE, sort=FALSE)
dt[is.na(method), method := "Unknown"]

# attach expected species + blank flag
dt <- merge(dt, expected_tbl, by="sample", all.x=TRUE, sort=FALSE)
dt[is.na(is_blank), is_blank := FALSE]  # unclassified -> not blank

# -------------------- observed species from alignment_genome --------------------
# default: unmapped if not aligned
dt[, observed_species := "unmapped"]
dt[aligned == TRUE, observed_species := "other_mapped"]

# map by prefix
# match: startsWith(alignment_genome, prefix) or grepl(prefix)
for (i in seq_len(nrow(mapping_tbl))) {
  pref <- mapping_tbl$prefix[i]
  sp   <- mapping_tbl$species[i]
  if (pref == "unmapped") next
  dt[aligned == TRUE & (startsWith(alignment_genome, pref) | grepl(pref, alignment_genome, fixed=TRUE)),
     observed_species := sp]
}
# If alignment_genome itself already looks like one of the species labels, keep it mapped
dt[aligned==TRUE & observed_species=="other_mapped" & alignment_genome %in% mapping_tbl$prefix,
   observed_species := alignment_genome]
```

```{r}
dt[, pass_mapq := TRUE]
if ("alignment_mapping_quality" %in% names(dt)) {
  dt[aligned == TRUE, pass_mapq := is.finite(alignment_mapping_quality) & alignment_mapping_quality >= MIN_MAPQ]
}

# core filters for comparisons
core <- dt[passes_filtering == TRUE]

# status for sample reads: correct vs misassigned vs unaligned/other
core[, status_sample := fifelse(
  sample %in% c("M","P","C","L","I","O"),

  fifelse(
    sample %in% c("I","O"),
    fifelse(
      observed_species == "unmapped",
      "unaligned",
      "misassigned"
    ),

    fifelse(
      observed_species == expected_species,
      "correct",
      fifelse(
        observed_species != "unmapped" &
          observed_species %in% expected_tbl[sample %in% c("M","P","C","L"),
                                             expected_species],
        "misassigned",
        fifelse(
          observed_species == "unmapped",
          "unaligned",
          "other_mapped"
        )
      )
    )
  ),

  NA_character_
)]

# group for unclassified barcodes (not in sample_sheet)
core[, status_unclassified := fifelse(sample == "unclassified",
                               fifelse(observed_species == "unmapped", "unclassified_unmapped", "unclassified_mapped"),
                               NA_character_)]
```

```{r}
# -------------------- Convenience aliases + grouping for downstream plots --------------------
core[, status := status_sample]
core[, pred_species := observed_species]

# Group samples for Water vs Genomic DNA plots used later
core[, truth_group := fifelse(sample %in% c("I","O"), "Water",
                       fifelse(sample %in% c("M","P","C","L"), "Genomic DNA", NA_character_))]

# Human-readable sample labels (Water + expected species names)
core[, sample_label := fifelse(truth_group == "Water", "Water", expected_species)]

# Tag input amount from experiment_id suffix
core[, input_ng := fifelse(grepl("3ng$", experiment_id), "3 ng",
                    fifelse(grepl("20ng$", experiment_id), "20 ng", NA_character_))]

core[experiment_id == "20251018_MPI_CLO", input_ng := "3 ng"] 

core[, input_ng := factor(input_ng, levels = c("3 ng", "20 ng"))]

```

### QC table: mapped read counts by method, sample, expected species, observed species, and status
```{r}
tab_sample <- core[sample %in% c("M","P","C","L","I","O") & pass_mapq == TRUE,
                   .N, by=.(method, sample, expected_species, observed_species, status_sample)]
fwrite(tab_sample, file.path(OUTDIR, "counts_sample_by_method_species_status.tsv"), sep="\t")
```

### misassignment rate by method

- Method-level misassignment fraction among mapped reads classified as correct or misassigned

```{r}
rate_by_method <- core[
  sample %in% c("M","P","C","L","I","O") &
  pass_mapq == TRUE &
  aligned == TRUE,
  .(
    n_mapped = sum(status_sample %in% c("correct","misassigned")),
    misassigned = sum(status_sample == "misassigned"),
    mis_frac = sum(status_sample == "misassigned") / sum(status_sample %in% c("correct","misassigned"))
  ),
  by = .(method)
]

setorder(rate_by_method, -mis_frac)
fwrite(rate_by_method, file.path(OUTDIR, "misassign_rate_by_method.tsv"), sep="\t")
```



### Table: barcode × mapping confusion summary

- Confusion matrix rows = barcode-linked samples; columns = target species plus Unmapped.
- Blank barcodes are retained in the confusion matrix so that reads assigned to blank barcodes but mapped to target species contribute to recall denominators.
- Per-barcode precision, recall, and F1 are computed for gDNA barcodes only.
- Final Table S4 reports the mean recall, precision, and F1 across all gDNA barcode instances within each method.


```{r}
gDNA_samples  <- c("M","P","C","L")
blank_samples <- c("I","O")

# =========================
# 1) Species columns to track
# =========================
target_species <- unique(na.omit(as.character(
  expected_tbl[sample %in% gDNA_samples, expected_species]
)))
sp_cols <- target_species
target_species_ext <- c(target_species, "Unmapped")

# =========================
# 2) Build confusion matrix (keep blanks in matrix)
#    Anything not mapped to target species is counted as "Unmapped"
# =========================
core_s4 <- core[
  method != "Unknown" &
    sample %in% c(gDNA_samples, blank_samples)
]

core_s4[, observed_species_s4 := ifelse(
  pass_mapq == TRUE & !is.na(observed_species) & observed_species %in% target_species,
  as.character(observed_species),
  "Unmapped"
)]

counts_long <- core_s4[
  , .(N = .N),
  by = .(experiment_id, method, sample, observed_species = observed_species_s4)
]

grid_rows <- unique(method_map[
  method != "Unknown" & sample %in% c(gDNA_samples, blank_samples),
  .(experiment_id = batch_dir, method, sample)
])

grid_full <- grid_rows[
  , .(observed_species = target_species_ext),
  by = .(experiment_id, method, sample)
]

s4_counts <- merge(
  grid_full, counts_long,
  by = c("experiment_id","method","sample","observed_species"),
  all.x = TRUE
)
s4_counts[is.na(N), N := 0L]

confusion_wide <- dcast(
  s4_counts,
  experiment_id + method + sample ~ observed_species,
  value.var = "N",
  fill = 0
)

fwrite(
  confusion_wide,
  file.path(OUTDIR, "confusion_barcode_by_species_S4_used.tsv"),
  sep = "\t"
)

# =========================
# 3) Per-gDNA-barcode precision / recall / F1
# =========================
exp_map <- unique(
  expected_tbl[sample %in% gDNA_samples, .(sample, expected_species)]
)

tab_g <- merge(
  confusion_wide[sample %in% gDNA_samples],
  exp_map,
  by = "sample",
  all.x = TRUE
)

# recall denominator includes blank rows mapped to target species
col_tot <- confusion_wide[
  , lapply(.SD, sum),
  by = .(experiment_id),
  .SDcols = sp_cols
]

col_tot_long <- melt(
  col_tot,
  id.vars = "experiment_id",
  variable.name = "expected_species",
  value.name = "col_total_expected"
)

tab_g[, row_total_species := rowSums(as.matrix(.SD)), .SDcols = sp_cols]

m_g   <- as.matrix(tab_g[, ..sp_cols])
idx_c <- match(tab_g$expected_species, sp_cols)

tp <- rep(NA_real_, nrow(tab_g))
ok <- which(!is.na(idx_c))
if (length(ok)) tp[ok] <- m_g[cbind(ok, idx_c[ok])]
tab_g[, tp := tp]

tab_g <- merge(
  tab_g,
  col_tot_long,
  by = c("experiment_id","expected_species"),
  all.x = TRUE
)

tab_g[, precision := fifelse(row_total_species > 0, tp / row_total_species, NA_real_)]
tab_g[, recall    := fifelse(col_total_expected > 0, tp / col_total_expected, NA_real_)]
tab_g[, f1 := fifelse(
  is.finite(precision) & is.finite(recall) & (precision + recall) > 0,
  2 * precision * recall / (precision + recall),
  NA_real_
)]

fwrite(
  tab_g[, .(
    experiment_id, method, sample, expected_species,
    tp, row_total_species, col_total_expected,
    precision, recall, f1
  )],
  file.path(OUTDIR, "Table_S4_barcode_instance_metrics.tsv"),
  sep = "\t"
)

# =========================
# 4) Final Table S4
# =========================
table_s4 <- tab_g[, .(
  Mean_Recall_pct    = 100 * mean(recall,    na.rm = TRUE),
  Mean_Precision_pct = 100 * mean(precision, na.rm = TRUE),
  Mean_F1_pct        = 100 * mean(f1,        na.rm = TRUE),
  n_gdna_instance    = .N,
  n_batches          = uniqueN(experiment_id)
), by = .(Method = method)]

setorder(table_s4, Method)

if (requireNamespace("knitr", quietly = TRUE)) {
  knitr::kable(table_s4, digits = 3)
} else {
  table_s4
}
```


### Summary: sample-level misassigned per million mapped reads with shared denominator

- target_species defined in the previous section is reused here

```{r}
gDNA_samples  <- c("M","P","C","L")
blank_samples <- c("I","O")
ord <- c("Standard", "SFB", "PLP", "SFB+PLP")
sample_label_levels <- c(
  "Water",
  "Mycoplasma hominis",
  "DCS_Lambda",
  "Phocaeicola vulgatus",
  "Corynebacterium striatum"
)

# Per-sample numerator and denominator pieces
sample_metrics <- core[
  method != "Unknown" &
    sample %in% c(gDNA_samples, blank_samples),
  .(
    total_mapped = sum(pass_mapq == TRUE & pred_species %in% target_species),
    tp = sum(
      pass_mapq == TRUE &
      pred_species %in% target_species &
      pred_species == expected_species
    )
  ),
  by = .(experiment_id, method, input_ng, sample, expected_species)
]

sample_metrics[, method := factor(method, levels = ord)]

sample_metrics[, sample_label := fifelse(
  sample %in% blank_samples,
  "Water",
  expected_species
)]

# For DNA samples: misassigned = target-mapped non-expected reads
# For blanks: misassigned = all target-mapped reads
sample_metrics[, misassigned_for_plot := fifelse(
  sample %in% blank_samples,
  total_mapped,
  pmax(total_mapped - tp, 0)
)]

# Shared denominator within each batch × method
denom_tbl <- sample_metrics[
  ,
  .(total_mapped_batch_method = sum(total_mapped, na.rm = TRUE)),
  by = .(experiment_id, method, input_ng)
]

plot_shared <- merge(
  sample_metrics,
  denom_tbl,
  by = c("experiment_id", "method", "input_ng"),
  all.x = TRUE,
  sort = FALSE
)

plot_shared[, misassigned_per_million_shared := fifelse(
  total_mapped_batch_method > 0,
  1e6 * misassigned_for_plot / total_mapped_batch_method,
  NA_real_
)]

pair_tbl <- unique(plot_shared[, .(experiment_id, method)])[
  ,
  .(batch_pair = paste(sort(unique(as.character(method))), collapse = " & ")),
  by = experiment_id
]

df_pairs <- merge(
  plot_shared[, .(
    experiment_id, method, input_ng,
    sample_label, misassigned_per_million_shared
  )],
  pair_tbl,
  by = "experiment_id",
  all.x = TRUE,
  sort = FALSE
)

df_pairs[, sample_label := factor(sample_label, levels = sample_label_levels)]
```

### Figure: Misassigned per million mapped reads by sample (shared denominator within batch × method)

```{r}
p2 <- ggplot(
  df_pairs[!is.na(batch_pair)],
  aes(x = sample_label, y = misassigned_per_million_shared, fill = method)
) +
  geom_col(position = position_dodge2(width = 0.8, preserve = "single")) +
  facet_grid(batch_pair ~ input_ng, scales = "free_y") +
  labs(
    x = "Sample",
    y = "Misassigned reads per million mapped reads\n(shared denominator within batch × method)",
    fill = "Method"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    text = element_text(size = 10),
    strip.background = element_rect(colour = NA)
  )
```

```{r fig.height=10,fig.width=8}
p2
```

### Figure: Mean sample-level misassigned per million mapped reads (Water vs Defined DNA)

```{r}
df_summary <- df_pairs[
  ,
  .(
    mean_mis = mean(misassigned_per_million_shared, na.rm = TRUE),
    sd_mis   = sd(misassigned_per_million_shared, na.rm = TRUE),
    n        = sum(!is.na(misassigned_per_million_shared))
  ),
  by = .(
    sample_group = fifelse(sample_label == "Water", "Water", "Genomic DNA"),
    method
  )
]

df_summary[, se_mis := sd_mis / sqrt(n)]
df_summary[, sample_group := factor(sample_group, levels = c("Water", "Genomic DNA"))]

p1 <- ggplot(df_summary, aes(x = sample_group, y = mean_mis, fill = method)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(
    aes(ymin = mean_mis - se_mis, ymax = mean_mis + se_mis),
    position = position_dodge(width = 0.8),
    width = 0.2
  ) +
  labs(
    x = NULL,
    y = "Misassigned reads per million mapped reads",
    fill = "Protocol"
  ) +
  scale_x_discrete(labels = c(
    "Genomic DNA" = "Defined DNA",
    "Water" = "Water blanks"
  )) +
  theme_bw() +
  theme(
    text = element_text(size = 12),
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )
```

```{r fig.height=5,fig.width=7}
p1
```



